---
title: "Model development"
author: "Anthony Yannarell"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose

The point of this document is to do data simulation to develop and validate statistical models -- particularly those for the kind of multilevel bayesian approach advocated by Richard McElreath. I have the split-plot design of the PHRE experiment particularly in mind, but there are some other things I've had kicking around that I'm throwing in here too.

I've been meaning to do this for a while, and I finally wimped out and started playing around with ChatGPT to develop the functions. It turned up some stuff that seems pretty useful, so I decided to put it all down here.

## Two-way ANOVA data

I started asking for just a function to generate two-way ANOVA with a specified design and effects structure.

```{r two-way anova, echo=FALSE}
# generated by ChatGPT

simulate_anova2 <- function(replications = 10,
                            levels_A = c("A1", "A2"),
                            levels_B = c("B1", "B2"),
                            effect_A = c(0, 1),
                            effect_B = c(0, 2),
                            interaction_effects = matrix(0, nrow = length(levels_A), ncol = length(levels_B)),
                            sd_error = 1,
                            grand_mean = 0,
                            seed = NULL) {
  
  if (!is.null(seed)) set.seed(seed)
  
  # Validate inputs
  stopifnot(length(effect_A) == length(levels_A),
            length(effect_B) == length(levels_B),
            all(dim(interaction_effects) == c(length(levels_A), length(levels_B))))
  
  # Create full design grid
  design <- expand.grid(A = levels_A, B = levels_B)
  
  # Generate the data
  sim_data <- do.call(rbind, lapply(1:nrow(design), function(i) {
    a_idx <- which(levels_A == design$A[i])
    b_idx <- which(levels_B == design$B[i])
    
    mu <- grand_mean + effect_A[a_idx] + effect_B[b_idx] + interaction_effects[a_idx, b_idx]
    y <- rnorm(replications, mean = mu, sd = sd_error)
    
    data.frame(
      A = rep(design$A[i], replications),
      B = rep(design$B[i], replications),
      Y = y
    )
  }))
  
  sim_data$A <- factor(sim_data$A, levels = levels_A)
  sim_data$B <- factor(sim_data$B, levels = levels_B)
  
  return(sim_data)
}

```

And here is an example of how to use this.

```{r two-way example}

# Define interaction effects as a matrix
interaction_matrix <- matrix(
  c(0,  1,
    -1, 0.5),
  nrow = 2, byrow = TRUE
)

# Simulate data
set.seed(123)
sim_df <- simulate_anova2(
  replications = 5,
  levels_A = c("Low", "High"),
  levels_B = c("Control", "Treatment"),
  effect_A = c(0, 3),           # A main effect: High = +3
  effect_B = c(0, -2),          # B main effect: Treatment = -2
  interaction_effects = interaction_matrix,
  sd_error = 1
)

# Preview the data
head(sim_df)

# Fitting the model in base R
anova_model <- aov(Y ~ A * B, data = sim_df)
summary(anova_model)
```

## Split-plot ANOVA data

Here is a function that will generate a basic split-plot dataset. There is a little extra decoration in that you can designate replication of the sub-plot factor within each (whole) plot, so it can be the basis of any kind of hierarchical nested design. Actually, upon studying this more, the sub-plot replication would be (technical) replication **within each** sub-plot unit. That is, not replicated sub-plots, but pseudoreplication within a subplot. So, basically, this is what Jacob did, when he collected two cores per PHRE plot.

Note that factor A is the whole-plot factor.

```{r split-plot anova, echo=FALSE}
# generated by ChatGPT

simulate_splitplot_anova2 <- function(wholeplot_reps = 2,
                                      subplot_reps = 3,
                                      levels_A = c("A1", "A2"),
                                      levels_B = c("B1", "B2"),
                                      effect_A = NULL,
                                      effect_B = NULL,
                                      interaction_effects = NULL,
                                      sd_wholeplot = 1,
                                      sd_subplot = 1,
                                      sd_residual = 1,
                                      grand_mean = 0,
                                      seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  
  n_A <- length(levels_A)
  n_B <- length(levels_B)
  
  # Set default effects if NULL
  if (is.null(effect_A)) effect_A <- rep(0, n_A)
  if (is.null(effect_B)) effect_B <- rep(0, n_B)
  if (is.null(interaction_effects)) interaction_effects <- matrix(0, nrow = n_A, ncol = n_B)
  
  stopifnot(length(effect_A) == n_A,
            length(effect_B) == n_B,
            all(dim(interaction_effects) == c(n_A, n_B)))
  
  sim_data <- data.frame()
  plot_id <- 1
  
  for (a_idx in seq_along(levels_A)) {
    for (wp_rep in seq_len(wholeplot_reps)) {
      # Simulate whole-plot error
      wholeplot_error <- rnorm(1, mean = 0, sd = sd_wholeplot)
      plot_label <- paste0("Plot_", plot_id)
      
      for (b_idx in seq_along(levels_B)) {
        # Simulate subplot error for this (A, B) combination within a plot
        subplot_error <- rnorm(1, mean = 0, sd = sd_subplot)
        
        mu <- grand_mean + effect_A[a_idx] + effect_B[b_idx] + interaction_effects[a_idx, b_idx]
        
        # Generate multiple measurements per subplot
        for (s in seq_len(subplot_reps)) {
          response <- rnorm(1, mean = mu + wholeplot_error + subplot_error, sd = sd_residual)
          
          sim_data <- rbind(sim_data, data.frame(
            A = levels_A[a_idx],
            B = levels_B[b_idx],
            Plot = plot_label,
            SubplotRep = s,
            Y = response
          ))
        }
      }
      plot_id <- plot_id + 1
    }
  }
  
  sim_data$A <- factor(sim_data$A, levels = levels_A)
  sim_data$B <- factor(sim_data$B, levels = levels_B)
  sim_data$Plot <- factor(sim_data$Plot)
  
  return(sim_data)
}
```

Here is an example of how to use this.

```{r split-plot example}

# Interaction matrix for 3x3 design
interaction_matrix <- matrix(
  c(0, 0.5, 1,
    -0.2, 0.3, 0.8,
    0.1, -0.4, 0),
  nrow = 3, byrow = TRUE
)

set.seed(123)
sim_df <- simulate_splitplot_anova2(
  wholeplot_reps = 2,
  subplot_reps = 4,
  levels_A = c("Low", "Med", "High"),
  levels_B = c("Fert1", "Fert2", "Fert3"),
  effect_A = c(0, 2, 4),
  effect_B = c(0, 1, -1),
  interaction_effects = interaction_matrix,
  sd_wholeplot = 2,
  sd_subplot = 1,
  sd_residual = 1.5,
  grand_mean = 10
)

head(sim_df)

# Fitting the model in base R
aov_model <- aov(Y ~ A * B + Error(Plot / B), data = sim_df)
summary(aov_model)

# Fitting the model with nlme
library(nlme)
lme_model <- lme(Y ~ A * B, random = ~1 | Plot/B, data = sim_df)
summary(lme_model)
```

## PHRE design

At this point, I couldn't help myself, so I decided go for a function that can simulate according to the actual PHRE design: a split-plot, where the sub-plot treatment structure is made of a 2-by-2-by-2 factorial. This version of the function allows for a full 4-way interaction matrix.

This is built off the previous function, so there can be multiple sub-plot reps, even though this isn't actually a thing in the PHRE design. And as above, factor A is the whole-plot factor (neonics, in the case of PHRE).

```{r PHRE-style ANOVA, echo=FALSE}
# Generated by ChatGPT

simulate_splitplot_4way <- function(wholeplot_reps = 2,
                                    subplot_reps = 3,
                                    levels_A = c("A1", "A2"),
                                    levels_B = c("B1", "B2"),
                                    levels_C = c("C1", "C2"),
                                    levels_D = c("D1", "D2"),
                                    effect_A = NULL,
                                    effect_B = NULL,
                                    effect_C = NULL,
                                    effect_D = NULL,
                                    interaction_ABCD = NULL,
                                    sd_wholeplot = 1,
                                    sd_subplot = 1,
                                    sd_residual = 1,
                                    grand_mean = 0,
                                    seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  
  n_A <- length(levels_A)
  n_B <- length(levels_B)
  n_C <- length(levels_C)
  n_D <- length(levels_D)
  
  # Default effects
  effect_A <- if (is.null(effect_A)) rep(0, n_A) else effect_A
  effect_B <- if (is.null(effect_B)) rep(0, n_B) else effect_B
  effect_C <- if (is.null(effect_C)) rep(0, n_C) else effect_C
  effect_D <- if (is.null(effect_D)) rep(0, n_D) else effect_D
  
  # Full 4-way interaction defaults to 0
  interaction_ABCD <- if (is.null(interaction_ABCD)) 
    array(0, dim = c(n_A, n_B, n_C, n_D)) else interaction_ABCD
  
  sim_data <- data.frame()
  plot_id <- 1
  
  for (a_idx in seq_along(levels_A)) {
    for (wp_rep in seq_len(wholeplot_reps)) {
      wholeplot_error <- rnorm(1, mean = 0, sd = sd_wholeplot)
      plot_label <- paste0("Plot_", plot_id)
      
      for (b_idx in seq_along(levels_B)) {
        for (c_idx in seq_along(levels_C)) {
          for (d_idx in seq_along(levels_D)) {
            mu <- grand_mean +
              effect_A[a_idx] +
              effect_B[b_idx] +
              effect_C[c_idx] +
              effect_D[d_idx] +
              interaction_ABCD[a_idx, b_idx, c_idx, d_idx]
            
            subplot_error <- rnorm(1, mean = 0, sd = sd_subplot)
            
            for (s in seq_len(subplot_reps)) {
              response <- rnorm(1, mean = mu + wholeplot_error + subplot_error, sd = sd_residual)
              
              sim_data <- rbind(sim_data, data.frame(
                A = levels_A[a_idx],
                B = levels_B[b_idx],
                C = levels_C[c_idx],
                D = levels_D[d_idx],
                Plot = plot_label,
                SubplotRep = s,
                Y = response
              ))
            }
          }
        }
      }
      plot_id <- plot_id + 1
    }
  }
  
  # Convert to factors
  sim_data$A <- factor(sim_data$A, levels = levels_A)
  sim_data$B <- factor(sim_data$B, levels = levels_B)
  sim_data$C <- factor(sim_data$C, levels = levels_C)
  sim_data$D <- factor(sim_data$D, levels = levels_D)
  sim_data$Plot <- factor(sim_data$Plot)
  
  return(sim_data)
}
```

Here's an example of usage. Note that in this example (and subsequent ones), the interaction matrix is generated randomly.

```{r PHRE-style example}

# Define A×B×C×D interaction effects (2x2x2x2 = 16 values)
abcd_effects <- array(runif(2*2*2*2, -1, 1), dim = c(2,2,2,2))

sim_df <- simulate_splitplot_4way(
  wholeplot_reps = 2,
  subplot_reps = 3,
  levels_A = c("Control", "Treatment"),
  levels_B = c("Low", "High"),
  levels_C = c("Short", "Long"),
  levels_D = c("Dry", "Wet"),
  effect_A = c(0, 3),
  effect_B = c(0, 1),
  effect_C = c(0, -1),
  effect_D = c(0, 2),
  interaction_ABCD = abcd_effects,
  sd_wholeplot = 1.2,
  sd_subplot = 0.8,
  sd_residual = 0.6,
  grand_mean = 10,
  seed = 42
)

head(sim_df)

# Fit the model in base R
aov_model <- aov(Y ~ A * B * C * D + Error(Plot / (B*C*D)), data = sim_df)
summary(aov_model)

# Fit the model with lme4
library(lme4)
lmer_model <- lmer(Y ~ A * B * C * D + (1 | Plot), data = sim_df)
summary(lmer_model)
```

## Helper functions and the full Monty

This stuff can be built on further. ChatGPT offered some additional functions to help with some of the effect sizes. This will generate a 4-by-4 array of (random) interaction effects. The main sparkle is that it will enforce a zero-sum constraint across all of the dimensions, and this will prevent any unintended shifts in the grand mean.

```{r zero-sum 4-by-4 interactions, echo=FALSE}
generate_effect_array_4way <- function(levels_A = c("A1", "A2"),
                                       levels_B = c("B1", "B2"),
                                       levels_C = c("C1", "C2"),
                                       levels_D = c("D1", "D2"),
                                       effect_size_range = c(-1, 1),
                                       seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  
  n_A <- length(levels_A)
  n_B <- length(levels_B)
  n_C <- length(levels_C)
  n_D <- length(levels_D)
  
  array_raw <- array(runif(n_A * n_B * n_C * n_D,
                           min = effect_size_range[1],
                           max = effect_size_range[2]),
                     dim = c(n_A, n_B, n_C, n_D))
  
  # Enforce sum-to-zero constraint over all dimensions
  array_adj <- array_raw
  array_mean <- mean(array_raw)
  array_adj <- array_adj - array_mean
  
  return(array_adj)
}

# Example of usage
abcd_effects <- generate_effect_array_4way(
  levels_A = c("Control", "Treatment"),
  levels_B = c("Low", "High"),
  levels_C = c("Short", "Long"),
  levels_D = c("Dry", "Wet"),
  effect_size_range = c(-1, 1),
  seed = 123
)

# Use it in simulation
sim_df <- simulate_splitplot_4way(
  wholeplot_reps = 2,
  subplot_reps = 2,
  interaction_ABCD = abcd_effects,
  effect_A = c(0, 2),
  effect_B = c(0, 1),
  effect_C = c(0, -1),
  effect_D = c(0, 0.5),
  grand_mean = 10,
  seed = 42
)

head(sim_df)
```

This helper function will also generate zero-sum interaction effects, but it gives more fine-scale control over lower-order effects.

It has the following defined inputs:

-   `levels_list`: a list of level vectors, e.g. `list(levels_A, levels_B)` or `list(levels_B, levels_C, levels_D)`

-   `effect_size_range`: numeric range for uniform random values

-   `sum_to_zero`: whether to subtract the grand mean from the array

-   `seed`: for reproducibility.

```{r zero-sum lower-order interactions, echo=FALSE}
generate_effect_array <- function(levels_list,
                                  effect_size_range = c(-1, 1),
                                  sum_to_zero = TRUE,
                                  seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  
  dims <- sapply(levels_list, length)
  effect_array <- array(runif(prod(dims),
                              min = effect_size_range[1],
                              max = effect_size_range[2]),
                        dim = dims)
  
  if (sum_to_zero) {
    grand_mean <- mean(effect_array)
    effect_array <- effect_array - grand_mean
  }
  
  return(effect_array)
}

# examples of usage

# 2-way B:C interaction
bc_effect <- generate_effect_array(
  levels_list = list(c("B1", "B2"), c("C1", "C2")),
  effect_size_range = c(-0.5, 0.5),
  seed = 101
)

# 3-wqy B:C:D interaction
bcd_effect <- generate_effect_array(
  levels_list = list(c("B1", "B2"), c("C1", "C2"), c("D1", "D2")),
  effect_size_range = c(-1, 1),
  seed = 42
)

# simulation
sim_df <- simulate_splitplot_4way(
  wholeplot_reps = 3,
  subplot_reps = 2,
  levels_A = c("Control", "Treatment"),
  levels_B = c("Low", "High"),
  levels_C = c("Short", "Long"),
  levels_D = c("Dry", "Wet"),
  effect_A = c(0, 1.5),
  effect_B = c(0, 0.5),
  effect_C = c(0, -0.5),
  effect_D = c(0, 1),
  interaction_ABCD = generate_effect_array(
    levels_list = list(c("Control", "Treatment"), c("Low", "High"), c("Short", "Long"), c("Dry", "Wet")),
    effect_size_range = c(-0.8, 0.8),
    seed = 100
  ),
  grand_mean = 12,
  seed = 1
)
```

And here is a function that will combine all of these specified effects into the full linear predictor model. ChatGPT says that this function can be used before simulation to preview the full expected mean structure, and that it will be useful to simulate residual variation around a known linear mean.

It uses these inputs:

-   `data`: A data frame with all factor combinations

-   `main_effects`: Named list of vectors, e.g. `list(A = c(0, 1.5))`

-   `interactions`: List of lists, each with:

    -   `$factors`: character vector of factor names

    -   `$effect`: multidimensional array from `generate_effect_array()`

-   `grand_mean`: the base intercept (μ₀).

```{r apply effects to linear predictor, echo=FALSE}
apply_effects <- function(data,
                          main_effects = list(),
                          interactions = list(),
                          grand_mean = 0) {
  mu <- rep(grand_mean, nrow(data))
  
  # Apply main effects
  for (factor_name in names(main_effects)) {
    if (!factor_name %in% names(data))
      stop(paste("Factor", factor_name, "not found in data"))
    levels_vec <- levels(data[[factor_name]])
    effect_vec <- main_effects[[factor_name]]
    if (length(effect_vec) != length(levels_vec))
      stop(paste("Length mismatch for main effect:", factor_name))
    mu <- mu + effect_vec[match(data[[factor_name]], levels_vec)]
  }
  
  # Apply interaction effects
  for (interaction in interactions) {
    factor_names <- interaction$factors
    effect_array <- interaction$effect
    
    # Get indices for each factor level
    idx_list <- lapply(factor_names, function(f) {
      if (!f %in% names(data))
        stop(paste("Factor", f, "not found in data"))
      match(data[[f]], levels(data[[f]]))
    })
    
    # Convert indices to matrix
    idx_mat <- do.call(cbind, idx_list)
    
    # Use mapply to extract appropriate element from effect_array
    mu <- mu + mapply(function(...) effect_array[...], 
                      as.data.frame(t(idx_mat)))
  }
  
  return(mu)
}

# Example
# Setup factors
levels_A <- c("Control", "Treatment")
levels_B <- c("Low", "High")
levels_C <- c("Short", "Long")
levels_D <- c("Dry", "Wet")

# Create full design
full_design <- expand.grid(
  A = factor(levels_A, levels_A),
  B = factor(levels_B, levels_B),
  C = factor(levels_C, levels_C),
  D = factor(levels_D, levels_D)
)

# Define effects
main_eff <- list(
  A = c(0, 2),
  B = c(0, 1),
  C = c(0, -1),
  D = c(0, 0.5)
)

interactions <- list(
  list(
    factors = c("A", "B", "C", "D"),
    effect = generate_effect_array(
      levels_list = list(levels_A, levels_B, levels_C, levels_D),
      effect_size_range = c(-0.5, 0.5),
      seed = 10
    )
  )
)

# Compute linear predictor
full_design$mu <- apply_effects(full_design, main_eff, interactions, grand_mean = 10)
head(full_design)
```

Finally, here is the full PHRE function that also embeds two of the helper functions defined above.

```{r PHRE data with helpers, echo=FALSE}
simulate_splitplot_full <- function(wholeplot_reps = 2,
                                    subplot_reps = 2,
                                    levels_A = c("A1", "A2"),
                                    levels_B = c("B1", "B2"),
                                    levels_C = c("C1", "C2"),
                                    levels_D = c("D1", "D2"),
                                    main_effects = list(),
                                    interactions = list(),
                                    grand_mean = 0,
                                    sd_wholeplot = 1,
                                    sd_subplot = 1,
                                    sd_residual = 1,
                                    seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  
  # Convert level names to factors
  A_levels <- factor(levels_A, levels_A)
  B_levels <- factor(levels_B, levels_B)
  C_levels <- factor(levels_C, levels_C)
  D_levels <- factor(levels_D, levels_D)
  
  # Full factorial for subplot combinations
  subplot_design <- expand.grid(
    A = A_levels,
    B = B_levels,
    C = C_levels,
    D = D_levels
  )
  
  # Compute deterministic μ values using effects
  mu_values <- apply_effects(
    data = subplot_design,
    main_effects = main_effects,
    interactions = interactions,
    grand_mean = grand_mean
  )
  subplot_design$mu <- mu_values
  
  sim_data <- data.frame()
  plot_id <- 1
  
  for (a_idx in seq_along(A_levels)) {
    A_level <- A_levels[a_idx]
    
    for (wp in seq_len(wholeplot_reps)) {
      plot_label <- paste0("Plot_", plot_id)
      wholeplot_error <- rnorm(1, 0, sd_wholeplot)
      
      for (row in seq_len(nrow(subplot_design))) {
        this_row <- subplot_design[row, ]
        if (this_row$A != A_level) next
        
        subplot_error <- rnorm(1, 0, sd_subplot)
        
        for (r in seq_len(subplot_reps)) {
          residual_error <- rnorm(1, 0, sd_residual)
          y <- this_row$mu + wholeplot_error + subplot_error + residual_error
          
          sim_data <- rbind(sim_data, data.frame(
            A = this_row$A,
            B = this_row$B,
            C = this_row$C,
            D = this_row$D,
            Plot = plot_label,
            SubplotRep = r,
            Y = y
          ))
        }
      }
      plot_id <- plot_id + 1
    }
  }
  
  sim_data$A <- factor(sim_data$A, levels = levels_A)
  sim_data$B <- factor(sim_data$B, levels = levels_B)
  sim_data$C <- factor(sim_data$C, levels = levels_C)
  sim_data$D <- factor(sim_data$D, levels = levels_D)
  sim_data$Plot <- factor(sim_data$Plot)
  
  return(sim_data)
}
```

Example of usage.

```{r example for combined simulation}
# Define levels
levels_A <- c("Control", "Treatment")
levels_B <- c("Low", "High")
levels_C <- c("Short", "Long")
levels_D <- c("Dry", "Wet")

# Define main effects
main_effects <- list(
  A = c(0, 2),
  B = c(0, 1),
  C = c(0, -1),
  D = c(0, 0.5)
)

# Define interactions
interactions <- list(
  list(
    factors = c("A", "B", "C", "D"),
    effect = generate_effect_array(
      levels_list = list(levels_A, levels_B, levels_C, levels_D),
      effect_size_range = c(-0.5, 0.5),
      seed = 123
    )
  ),
  list(
    factors = c("B", "D"),
    effect = generate_effect_array(
      levels_list = list(levels_B, levels_D),
      effect_size_range = c(-0.3, 0.3),
      seed = 234
    )
  )
)

# Simulate data
sim_df <- simulate_splitplot_full(
  wholeplot_reps = 2,
  subplot_reps = 3,
  levels_A = levels_A,
  levels_B = levels_B,
  levels_C = levels_C,
  levels_D = levels_D,
  main_effects = main_effects,
  interactions = interactions,
  grand_mean = 10,
  sd_wholeplot = 1,
  sd_subplot = 0.8,
  sd_residual = 0.5,
  seed = 42
)

head(sim_df)
```
